A general and efficient representation of Ancestral Recombination Graphs

### 2022-08-25 paper.tex
More text on the continuum of recombination information

\newcommand{\noderef}[1]{\textsf{#1}}
% \newcommand{\msprime}[0]{\texttt{msprime}}

@@

\section*{Precision of recombination information}
Ancestral recombination graphs are often characterised
as consisting of a complete record of the genealogical history
of a sample. For example,
\cite{rasmussen2014genome} state that an ``ARG provides a record of all
coalescence and recombination events since the divergence of the sequences
under study'' and
according to \cite{deng2021distribution} an ARG
``provides all the information about the genealogical history of a sample,
including the locations of recombination events.'' It is worth
questioning, however, whether such complete and precise information is
realistically available, and in particular whether our
\emph{representation} of genealogical history should require
such precision.

@@

\begin{figure} \begin{center}
\includegraphics[width=\textwidth]{illustrations/inference.pdf} \end{center}
\caption{\label{fig-inferred-args} ARGs inferred by different methods from a
2.7kb region of the \textit{Drosophila melanogaster} ADH locus for 11
samples~\citep{kreitman1983nucleotide}. Note that the position of points on the
Y axis is that chosen for graphical clarity by the graphviz \texttt{dot}
algorithm,  rather than being
a direct measure of time. Additional notes about these ARGs.
}
\end{figure}

To illustrate this critical point, we inferred ARGs for the
the \citet{kreitman1983nucleotide} \emph{Drosophila melanogaster} dataset
(classically used as a benchmark for ARG inference),
using four recent inference methods and show the results in
Fig~\ref{fig-inferred-args}).

There is some consistency among the four inferred ARGs (such as the samples
FR-F, WA-F and Af-F being grouped into the same clade), but also considerable
differences in the order and number of events. \texttt{KwARG}
\citep{ignatieva2021kwarg} produces a parsimonious ARG with seven recombination
breakpoints, which has been shown to be the true minimum required to explain
the dataset when disallowing recurrent mutation \citep{song2003parsimonious};
this is a lower bound that, in general, significantly underestimates the actual
number of recombinations that might have occurred. An ARG drawn from the
posterior sample produced by \texttt{ARGweaver} has 17 breakpoints. Both tools
explicitly record the sequence undergoing each recombination event, and the
precise location of the breakpoint along the genome. \texttt{Relate}
\citep{speidel2019method} and \texttt{tsinfer} \citep{kelleher2019inferring} do
not place recombination events onto the ARG topology, therefore the transitions
between local trees are not explicitly inferred in terms of SPR operations.
\texttt{Relate} first infers local tree topologies and then identifies edges
that persist from one tree to the next as a separate step, while
\texttt{tsinfer} explicitly constructs nodes that are shared across local
trees. The number of inferred trees also depends on the parameter settings; all
four tools can allow for the presence of sequencing errors and recurrent
mutations, reducing the number of inferred local trees (in this case, based on
the input mutation and recombination rate \texttt{Relate} hence posits only two
recombination breakpoints).

% Note that all four methods produce output which can be represented in the
% succinct tree sequence format, and which is consistent with our definition of
% an ARG as a data structure. The ARGs produced by \texttt{KwARG} and
% \texttt{ARGweaver} contain more nodes to record the recombination events
% (corresponding to Figure 4 stage `B'), while the ARGs output by
% \texttt{tsinfer} and \texttt{Relate} being, in essence, simplified (roughly
% corresponding to Figure 4 stages `C' and 'D', respectively).

>>>>

\newcommand{\noderef}[1]{\textsf{#1}}
\newcommand{\tsinfer}[0]{\texttt{tsinfer}}
\newcommand{\kwarg}[0]{\texttt{KwARG}}
\newcommand{\argweaver}[0]{\texttt{ARGweaver}}
\newcommand{\relate}[0]{\texttt{Relate}}
\newcommand{\espalier}[0]{\texttt{Espalier}}

@@

\section*{Precision of ancestry information}
Ancestral recombination graphs are often characterised
as consisting of a complete record of the genealogical history
of a sample. For example,
\cite{rasmussen2014genome} state that an ``ARG provides a record of all
coalescence and recombination events since the divergence of the sequences
under study'' and
according to \cite{deng2021distribution} an ARG
``provides all the information about the genealogical history of a sample,
including the locations of recombination events.'' It is worth
questioning, however, whether such comprehensive information is
realistically available, and in particular whether our
representation of genealogical history should \emph{require}
such precision.

@@

\begin{figure} \begin{center}
\includegraphics[width=\textwidth]{illustrations/inference.pdf} \end{center}
\caption{\label{fig-inferred-args} ARGs inferred by
(A) \kwarg~\citep{ignatieva2021kwarg},
(B) \argweaver~\citep{rasmussen2014genome,hubisz2020inference},
(C) \tsinfer~\citep{kelleher2019inferring},
and (D) \relate~\citep{speidel2019method}
for 11 samples of a 2.7kb region of the \textit{Drosophila melanogaster} ADH
locus~\citep{kreitman1983nucleotide}. The \argweaver\ example
is a sample from the inferred posterior distribution.
Note that the position of points on the
Y axis is that chosen for graphical clarity by the graphviz \texttt{dot}
algorithm,  rather than being
a direct measure of time. Additional notes about these ARGs.
}
\end{figure}

To illustrate this critical point, we inferred ARGs for the
the classical \citet{kreitman1983nucleotide} benchmark dataset
using four recent inference methods and show the results in
Fig~\ref{fig-inferred-args}.
While there is some consistency among the inferred ARGs (such as the samples
FR-F, WA-F and Af-F being grouped into the same clade), there
are also considerable differences. For this dataset, the minimum number
of recombination breakpoints required in the absence of recurrent mutation
is 7~\citep{song2003parsimonious}, and the number of inferred breakpoints
under the input mutation and recombination rates
varies from 2 (\relate) to 17 (\argweaver).
The ARGs inferred by the different methods also have major differences
in how recombination information is represented. Essentially,
\kwarg\ and \argweaver\ infer a specific causal recombination \emph{event}
for each breakpoint, and those events are fully
described in the output graph. On the other hand, \tsinfer\ and
\relate\ do not infer causal events but instead focus on
representing the inferred \emph{effect}: breakpoints between the
trees. It is important to note that this does not imply that
the methods must then necessarily infer only a sequence of
unrelated local trees
% More? See https://github.com/tskit-dev/what-is-an-arg-paper/issues/38
% for discussion. I'm sure there were other examples of people saying
% that
(as claimed by e.g.~\citep{hejase2020summary}),
and information about recombination and the correlation structure
between trees is not ``all or nothing''.
Both methods capture correlation structure between trees:
\relate\ first infers local tree topologies and then identifies edges
that persist from one tree to the next as a separate step
(in a similar manner to \espalier~\citep{rasmussen2022espalier}), while
\texttt{tsinfer} explicitly constructs nodes that are shared across local
trees.

Because the ARGs inferred by \tsinfer\ and \relate\ do not contain
explicit recombination events we cannot compute a likelihood
under the coalescent with recombination (see appendix XXX).
[Some stuff about whether we should see this as necessary or not,
but maybe we should keep this for the conclusion?]
